---
title: "Equality in access to and participation in parkrun events in England over time"
output: html_notebook
date: 2020-06-05
---

```{r include=FALSE}
options(scipen = 99)
```

<br>
**Robert Smith**<sup>1</sup>    
**Paul Schneider**<sup>1</sup>    

<sup>1</sup>*ScHARR, University of Sheffield*       

****

#### Objectives of this analysis:
 1. Assess how access to and participation in parkrun events in England evolved over time
 2. Study the socioeconomic gradient in access and participation over time

****

## ANALYSIS

#### Overview

  We combine data on community deprivation with information on access to parkrun events and participation therein, and assess the the socioeconomic gradient in access and participation using the slope index of inequality. We then study trends in these characteristics over time.

  All analyses are conducted on the level of **lower layer super output areas (LSOAs)** are census areas define dy the English National Office of Statistics. There are 32,844 LSOAs in England. On average, they contain a population of about 1,700 individuals.  
  

#### Preparation

First of all, we clean the R environment and install, if necessary, and load the packages that are required for the analysis.

```{r}
# clean the R enviroment
rm(list=ls())
```

```{r message=FALSE, warning=FALSE}
# Load the required packages
library(dplyr)
library(reshape2)
library(data.table)
# library(date)
library(raster)
library(geosphere)
library(ggplot2)
library(scales)
library(RColorBrewer)
```

#### Data sets

The following data sets are used for the analysis:

  1. parkrun participation: the dataset contains information on the number of individuals from all Lower-Layer Super Output areas (LSOAs) who finished a parkun in any given week from January 1st 2010 to December 28th 2019.

  2. Parkrun event locations: the location of parkrun courses and the date when they were established.

  3. The 2015 Index of multiple deprivation for England: this data set provides a score on the relative deprivation level for each of the 32,8444 LSOAs in England. 
  
  4. ONS LSOA data: Population-weighted LSOA centroids and total population






##### Read in the data

Read in the raw data, clean and prepare 
 
```{r}
# 1. Parkrun participation data
  # read in parkrun participation data 
  runs_df = readRDS("./runs_per_lsoa_2010to2020.Rds")
  # participation is measured in n of idividuals who finish a parkrun event
  head(runs_df) # take a look at the data set
  dim(runs_df) # it is quite huge
  runs_df = as.data.table(runs_df) # we use data.table to speed up the analysis 
  runs_df$date = as.Date(unclass(runs_df$date),format="%Y-%m-%d") # convert to Date class
  
  # Clean 
  runs_df = runs_df[grep(pattern = "E",runs_df$lsoa),] # select LSOAs in England
  runs_df = runs_df[runs_df$date <= as.Date("2018-12-12"),] # select runs up until 2018-12-12
  
  # We will analyse the data on the monthly level, so 
  # we need to create  a year+month variable:   
  runs_df$month_year = substr(runs_df$date,1,7)
  head(runs_df)

# 2. parkrun event locations
  # read in data
  event_locations = read.csv("./event_info_20181212.csv")
  event_locations$month_year = substr(event_locations$Estblsh,1,7)
  event_locations$Estblsh = as.Date(event_locations$Estblsh)
  head(event_locations[,1:4])
  
  
  
# 3. 2015 IMD data  
  # Read in IMD scores
  lsoa_imd = read.csv("./IMD_data.csv")
  head(lsoa_imd) # take a look at the raw data
  
  # remove columsn we dont nee and rename the one we do need
  lsoa_imd = lsoa_imd %>% # read csv
    dplyr::select(c("LSOA.code..2011.","Index.of.Multiple.Deprivation..IMD..Score", # select relevant colums
             "Index.of.Multiple.Deprivation..IMD..Decile..where.1.is.most.deprived.10..of.LSOAs.",
             "Total.population..mid.2012..excluding.prisoners.")) %>%
    'colnames<-'(c("lsoa","imd_score","imd_decile","total_pop")) # rename columns
  head(lsoa_imd) # take a look at the clean data set

  
# 4. LSOA pop-weighted centroid locations
  # read in shape file
  lsoa_locations = shapefile("./England_lsoa_2011_centroids/england_lsoa_2011_centroids.shp")
  # use correct projection
  lsoa_locations = spTransform(lsoa_locations,crs("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
  # extract coordinates 
  lsoa_centr = coordinates(lsoa_locations)
  rownames(lsoa_centr) = lsoa_locations$code
  
```


#### Variables of interest


1. **parkrun participatin**: finishers...

2. **Distance to the nearest event**: The access of the population living in a given LSOA is measured as the geodesic distance (this is distance as the crow flies) between the LSOA's population-weighted centroid and the nearest active parkrun event.

3. **Index of multiple deprivation (IMD)**: The score combines indices from seven domains (health, education, barriers to horsing and services, crime and living environment), Higher score indicate higher levels of deprivation. For the analysis, we will mainly use IMD deciles: i.e. look at the top and bottom 10% LSOAs.


4. **Slope index of inequality**: used for access and participation


```{r}
# 1. parkrun participation

  # aggregate by lsoa and year (fast version using data.table)
  monthly_runs = runs_df[, lapply(.SD, sum), by = month_year, .SDcols = "finishers"]

  monthly_runs_by_lsoa = runs_df[, lapply(.SD, sum), by = c("month_year","lsoa"), .SDcols = "finishers"]
  
# Before we proceed with the analysis of socio economic disparities, 
  # we need to account for the missing data 
  # LSOAs with no runs are not in the data 
  # (some lsoas might even be missing completely, if they never had a participant)
  # so we have to fill them in manually:
  fill_dat = expand.grid(unique(monthly_runs_by_lsoa$month_year),
                         lsoa_imd$lsoa) # using all English LSOAs from the imd data set
  names(fill_dat) = c("month_year","lsoa")
  fill_dat = data.table(fill_dat)
  monthly_runs_by_lsoa = merge(fill_dat,monthly_runs_by_lsoa,by=c("month_year","lsoa"),all.x=T)
  monthly_runs_by_lsoa$finishers[is.na(monthly_runs_by_lsoa$finishers)] = 0  # filling missings
  dim(monthly_runs_by_lsoa) # now we have quite a big data set
  head(monthly_runs_by_lsoa)
  
  # proportion of LSOAs that had 0 runs in any given months
  monthly_runs_by_lsoa[, lapply(.SD, function(x){round(sum(x==0)/length(x),4)*100}), by = c("month_year"), .SDcols = "finishers"]
  
  
# 2. Access (= distance to the nearst parkrun event)
  # compute the distances between all 32,844 lsoas and all 465 parkrun events 
  # CAVE: computationally heavy computation!
  distM = geosphere::distm(x= lsoa_locations,y=cbind(event_locations$lng,event_locations$lat))
  rownames(distM) = rownames(lsoa_centr)
  colnames(distM) = event_locations$course
  # custom function to determine the distance to the nearest parkrun event
  # for any given month
  get_min_dist = function(available_event_cols = T,distance_matrix  = distM,month_year="2010-01"){
     Mat = distance_matrix[,available_event_cols]
     min_dist = apply(Mat,1,min)
     min_dist_df = data.frame(lsoa = rownames(Mat),month_year=month_year,access=min_dist)
     return(min_dist_df)
  }
  
  
  ## loop to assess access in any given month  (month_years)
  # (may take a while to run)
  distance.df = c()
  months_t = as.Date(paste(unique(monthly_runs_by_lsoa$month_year),"-01",sep=""))
  months_t = seq(from=min(months_t),max(months_t) ,by="months")+14 # for simplicity we look at middle of the month
    
  for(t in months_t){
      t = as.Date(t, origin = as.Date("1970-01-01"))
      # cat("\r  at:",as.character(t))
      events_in_operation = event_locations$Estblsh <= t # which parkrun events were active in month t?
      temp.dist = get_min_dist(available_event_cols = events_in_operation,month_year = substr(t,1,7))
      distance.df = rbind(distance.df,temp.dist)
    }
  
  head(distance.df)
  dim(distance.df)
  
  # merge
  distance.df = data.table(distance.df)
  monthly_runs_by_lsoa = merge(monthly_runs_by_lsoa,distance.df,by=c("lsoa","month_year"),all.x=T) 
  
  
# 3. now we cmobine the dat set with information on LSOA IMD deciles
  lsoa_imd = data.table(lsoa_imd)
  monthly_runs_by_lsoa = merge(monthly_runs_by_lsoa,lsoa_imd,by="lsoa")
  # length(unique(monthly_runs_by_lsoa$lsoa))
  
# 4. the SII will be computed later on - see below
```

****

## RESULTS

### Q1: Access to and participation in parkrun events in England over time

#### Access over time
```{r}
# median distance to the nearets event over time
access_ts = monthly_runs_by_lsoa[, lapply(.SD, median), by = month_year, .SDcols = "access"]
access_ts$month_year = as.Date(paste(access_ts$month_year,"-01",sep=""))
ggplot(access_ts,aes(x=month_year,y=access)) +
  geom_point() +
  geom_line() +
  ylim(c(0,40000)) +
  ylab("Median distance to the nearest parkrun event")

# % population living within 5km of a parkrun event
access_prop_ts = monthly_runs_by_lsoa[, lapply(.SD, function(x){sum(x<=5000)/length(x)}), by = month_year, .SDcols = "access"]
access_prop_ts$month_year = as.Date(paste(access_prop_ts$month_year,"-01",sep=""))
ggplot(access_prop_ts,aes(x=month_year,y=access*100)) +
  geom_point() +
  geom_line() +
  ylim(c(0,100)) +
  ylab("% population living within 5km of a parkrun event")
  
```


#### Participation over time
```{r}
# total participants per month across England
runs_ts_sum = monthly_runs_by_lsoa[, lapply(.SD, sum), by = month_year, .SDcols = "finishers"]
runs_ts_sum$month_year = as.Date(paste(runs_ts_sum$month_year,"-01",sep=""))
ggplot(runs_ts_sum,aes(x=month_year,y=finishers)) +
  geom_point() +
  geom_line() +
  ylab("total number of participants per month")

# median participants per LSOA per month
runs_ts_med = monthly_runs_by_lsoa[, lapply(.SD, median), by = month_year, .SDcols = "finishers"]
runs_ts_med$month_year = as.Date(paste(runs_ts_med$month_year,"-01",sep=""))
ggplot(runs_ts_med,aes(x=month_year,y=finishers)) +
  geom_point() +
  geom_line() +
  ylab("median participants per LSOA per month")
  
  # total participants per month across England
runs_ts_zero = monthly_runs_by_lsoa[, lapply(.SD, function(x){sum(x==0)/length(x)}), by = month_year, .SDcols = "finishers"]
runs_ts_zero$month_year = as.Date(paste(runs_ts_zero$month_year,"-01",sep=""))
ggplot(runs_ts_zero,aes(x=month_year,y=finishers)) +
  geom_point() +
  geom_line()+ 
  ylab("Proportion of LSOAs with zero participants per month") +
  ylim(c(0,1))
  
```



<br><br>

****

### Q2: The socioeconomic gradient in access and participation over time


#### The socioeconomic gradient in access over time

```{r}
access_by_imd = monthly_runs_by_lsoa[, lapply(.SD, median), by = c("month_year","imd_decile"), .SDcols = "access"]
access_by_imd$month_year = as.Date(paste(access_by_imd$month_year,"-01",sep=""))

imd_colors = brewer.pal(10, "RdYlGn")
imd_labels = c(c("Most deprived",paste(9:6,"th decile",sep=""),"Median",paste(4:2,"th decile",sep=""),"Least deprived"))

# median distance to the nearest event by imd decile
ggplot(access_by_imd,aes(x=month_year,y=access,col=as.factor(imd_decile),group=imd_decile)) +
  # geom_point() +
  geom_line() +
  scale_color_manual(values = imd_colors,name="IMD decile",labels=imd_labels) +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylim(c(0,50000)) +
  ylab("Median distance to the nearest parkrun event") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))

# Compute the ratio between the top and bottom IMD median distance ratio
access_by_imd_tb = dcast(data = access_by_imd,formula = month_year ~ imd_decile) %>%
  mutate(RII = `10` / `1`) %>%
    dplyr::select(month_year, RII) 

ggplot(access_by_imd_tb,aes(x=month_year,y=RII)) +
  geom_point() +
  geom_line() +
  # scale_color_manual(values = imd_colors,name="IMD decile",labels=imd_labels) +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylab("Ratio between top and bottom IMD median distance to nearest event") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))

```


#### The slope index of inequality in access over time

```{r}
# max and min IMD
imd_range =  c(max(monthly_runs_by_lsoa$imd_score),min(monthly_runs_by_lsoa$imd_score))

# Slope index of inequality - Example
temp = monthly_runs_by_lsoa[month_year=="2018-08", ]
lm.temp = lm(access ~ imd_score,temp)
pred.temp = predict(lm.temp,newdata = data.frame(imd_score = imd_range))
tb_ratio.temp = pred.temp[2] / pred.temp[1]
tb_ratio.temp 

ggplot() +
  geom_point(data=temp,aes(x=imd_score,y=access),alpha=0.5,size=0.5,col="cadetblue4") +
  geom_smooth(data=temp,aes(x=imd_score,y=access),col="darkorange",method="lm",se=F) +
  geom_point(aes(x=imd_range,y=pred.temp),size=3,col="red") +
  ylab("Distance from LSOA centroid to the nearest parkrun event") +
  xlab("LSOA IMD score") +
  ggtitle("Relationship between IMD and access in August 2018 - Slope of Inequality") +
  theme_minimal()

# Loop for all months
uniq.months = unique(monthly_runs_by_lsoa$month_year)
si_access_ts = c()
for(m in uniq.months){
  # cat("\r",m)
  temp = monthly_runs_by_lsoa[month_year==m, ]
  lm.temp = lm(access ~ imd_score,temp)
  pred.temp = predict(lm.temp,newdata = data.frame(imd_score = imd_range))
  tb_ratio.temp = pred.temp[2] / pred.temp[1]
  res.df.temp = data.frame(sii = tb_ratio.temp,month_year=m)
  si_access_ts = rbind(si_access_ts,res.df.temp)
}

si_access_ts$month_year = as.Date(paste(si_access_ts$month_year,"-01",sep=""))
ggplot(si_access_ts,aes(x=month_year,y=sii)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylab("Distance to the nearest parkrun event - SII") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))


```

__Conclusion: bottom has gained better access over time, only from 2017 onwards, inequality started to decrease__

<br><br>

****

#### The socioeconomic gradient in participation over time

```{r}
# total (sum) participation by imd group
runs_by_imd = monthly_runs_by_lsoa[, lapply(.SD, sum), by = c("month_year","imd_decile"), .SDcols = "finishers"]
runs_by_imd$month_year = as.Date(paste(runs_by_imd$month_year,"-01",sep=""))

imd_colors = brewer.pal(10, "RdYlGn")
imd_labels = c(c("Most deprived",paste(9:6,"th decile",sep=""),"Median",paste(4:2,"th decile",sep=""),"Least deprived"))

# median distance to the nearest event by imd decile
ggplot(runs_by_imd,aes(x=month_year,y=finishers,col=as.factor(imd_decile),group=imd_decile)) +
  # geom_point() +
  geom_line(size=0.5,alpha=0.5) +
  geom_smooth(se=F) +
  scale_color_manual(values = imd_colors,name="IMD decile",labels=imd_labels) +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylab("Total runs per month by imd (smoothed curves)") +
  xlab("Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))

# Compute the ratio between the top and bottom IMD median distance ratio
runs_by_imd_shares = dcast(data = runs_by_imd,formula = month_year ~ imd_decile,value.var = "finishers") 
runs_by_imd_shares$totals = rowSums(runs_by_imd_shares[,-1])
runs_by_imd_shares2 = runs_by_imd_shares[, lapply(.SD, function(x){x/totals}), .SDcols = 2:11]
runs_by_imd_shares2$month_year = runs_by_imd_shares$month_year
runs_by_imd_shares2 = melt(runs_by_imd_shares2,id.vars = "month_year")

ggplot(runs_by_imd_shares2,aes(x=month_year,y=value,fill=variable)) +
  geom_area(col="black",size=0.2,alpha=0.75) +
  scale_fill_manual(values = imd_colors,name="IMD decile",labels=imd_labels) +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylab("Share in total runs per month by IMD deciles") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))

```


#### The slope index of inequality in participation over time

```{r}
# max and min IMD
imd_range =  c(max(monthly_runs_by_lsoa$imd_score),min(monthly_runs_by_lsoa$imd_score))
total_pop_mean = mean(monthly_runs_by_lsoa$total_pop)

# Slope index of inequality - Example
temp = monthly_runs_by_lsoa[month_year=="2018-08", ]
lm.temp = glm(finishers ~ imd_score,data=temp,family=poisson(link="log"),offset = log(total_pop))
pred.temp = predict(lm.temp,newdata = data.frame(imd_score = imd_range,total_pop=total_pop_mean),type="response")
pred.temp.line = predict(lm.temp,
                         newdata = data.frame(imd_score = seq(imd_range[1],imd_range[2],by=-1),
                                              total_pop=total_pop_mean),
                         type="response")
tb_ratio.temp = pred.temp[2] / pred.temp[1]
tb_ratio.temp 

ggplot() +
  geom_point(data=temp,aes(x=imd_score,y=finishers),alpha=0.5,size=0.5,col="cadetblue4") +
  geom_line(aes(x=seq(imd_range[1],imd_range[2],by=-1),y=pred.temp.line),size=1,col="darkorange") +
  # geom_smooth(data=temp,aes(x=imd_score,y=finishers),col="darkorange",method="lm",se=F) +
  geom_point(aes(x=imd_range,y=pred.temp),size=3,col="red") +
  ylab("Distance from LSOA centroid to the nearest parkrun event") +
  xlab("LSOA IMD score") +
  ggtitle("Relationship between IMD and access in August 2018 - Slope of Inequality") +
  theme_minimal()


# Loop over all months
uniq.months = unique(monthly_runs_by_lsoa$month_year)
si_runs_ts = c()
for(m in uniq.months){
  # cat("\r",m)
  temp = monthly_runs_by_lsoa[month_year==m, ]
  lm.temp = glm(finishers ~ imd_score,data=temp,family=poisson(link="log"),offset = log(total_pop))
  pred.temp = predict(lm.temp,newdata = data.frame(imd_score = imd_range,total_pop=total_pop_mean),type="response")
  tb_ratio.temp = pred.temp[2] / pred.temp[1]
  res.df.temp = data.frame(sii = tb_ratio.temp,month_year=m)
  si_runs_ts = rbind(si_runs_ts,res.df.temp)
}

si_runs_ts$month_year = as.Date(paste(si_runs_ts$month_year,"-01",sep=""))

ggplot(si_runs_ts,aes(x=month_year,y=sii)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks = "1 year",date_labels = "%Y") +
  ylab("Monthly parkrun participation - SII") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=35))


```

__Conclusion: Inital decrease in inequality (2010-2013), then stable plateau around 35(!) __



<br><br>

****

## CONCLUSION

...


